\documentclass[9pt, twocolumn, oneside, a4paper]{memoir}
\usepackage{fontspec}
\setmainfont{Alegreya}
\usepackage{ragged2e}
\newfontfamily\scshape{Alegreya SC}
\newfontfamily\headingfont{Alegreya Sans SC}
\newfontfamily\subheadingfont[
    BoldFont={Alegreya Sans Medium},
    ItalicFont={Alegreya Sans Light Italic},
    BoldItalicFont={Alegreya Sans Medium Italic}
]{Alegreya Sans Light}
\newfontfamily\greekfont[Script=Greek, Scale=1.14, WordSpace=0.7]{GFS Neohellenic}
\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage[variant=ancient]{greek}
\newcommand{\greektext}[1]{\foreignlanguage{greek}{#1}}

\usepackage{microtype}
\emergencystretch \hsize
\tolerance 300%
\setlength{\parskip}{0pt}
\frenchspacing

\usepackage{titlesec}
\titleformat*{\section}{\headingfont}

\setlrmarginsandblock{1.37in}{*}{*}
\setulmarginsandblock{1.29in}{*}{2}

\checkandfixthelayout
\setmarginnotes{\columnsep}{0.93in}{0.5\onelineskip}

\pretitle{\par\raggedright\Huge\subheadingfont}
\title{\textit{Against Heresies}, ii.10; iii.1.1-2.3, 6.1.2, 16.1-6, 18.1-7, 19.1-3; iv.20.1-4}
\posttitle{ \textperiodcentered} 
\preauthor{\raggedright\Huge\subheadingfont}
\author{ Irenaeus of Lyons}
\postauthor{\vskip 0.5em}
\predate{}
\date{}
\postdate{}

\newcommand{\gloss}[1]{%
    \marginpar[\RaggedLeft \footnotesize{#1}]{\RaggedRight \footnotesize{#1}}}

\begin{document}
\maketitle
\section*{Perverse interpretations of Scripture by the heretics: God created all things out of nothing, and not from pre-existent matter.}


1. It is therefore in the highest degree irrational, that we should take no account of Him who is truly God, and who receives testimony from all, while we inquire whether there is above Him that [other being] who really has no existence, and has never been proclaimed by any one. For that nothing has been clearly spoken regarding Him, they themselves furnish testimony; for since they, with wretched success, transfer to that being who has been conceived of by them, those parables [of Scripture] which, whatever the form in which they have been spoken, are sought after [for this purpose], it is manifest that they now generate another [god], who was  never previously sought after. For by the fact that they thus endeavour to explain ambiguous passages of Scripture (ambiguous, however, not as if referring to another god, but as regards the dispensations of [the true] God), they have constructed another god, weaving, as I said before, ropes of sand, and affixing a more important to a less important question. For no question can be solved by means of another which itself awaits solution; nor, in the opinion of those possessed of sense, can an ambiguity be explained by means of another ambiguity, or enigmas by means of another greater enigma, but things of such character receive their solution from those which are manifest, and consistent and clear.

2. But these [heretics], while striving to explain passages of Scripture and parables, bring forward another more important, and indeed impious question, to this effect, ``Whether there be really another god above that God who was the Creator of the world?'' They are not in the way of solving the questions [which they propose]; for how could they find means of doing so? But they append an important question to one of less consequence, and thus insert [in their speculations] a difficulty incapable of solution. For in order that they may  know ``knowledge'' itself (yet not learning this fact, that the Lord, when thirty years old, came to the baptism of truth), they do impiously despise that God who was the Creator, and who sent Him for the salvation of men. And that they may be deemed capable of informing us whence is the substance of matter, while they believe not that God, according to His pleasure, in the exercise of His own will and power, formed all things (so that those things which now are should have an existence) out of what did not previously exist, they have collected [a multitude of] vain discourses. They thus truly reveal their infidelity; they do not believe in that which really exists, and they have fallen away into [the belief of] that which has, in fact, no existence.

3. For, when they tell us that all moist substance proceeded from the tears of Achamoth, all lucid substance from her smile, all solid substance from her sadness, all mobile substance from her terror, and that thus they have sublime knowledge on account of which they are superior to others—how can these things fail to be regarded as worthy of contempt, and truly ridiculous? They do not believe that God (being powerful, and rich in all resources) created matter itself, inasmuch as they know not how much a spiritual and divine essence can accomplish. But they do believe that their Mother, whom they style a female from a female, produced from her passions aforesaid the so vast material substance of creation. They inquire, too, whence the substance of creation was supplied to the Creator; but they do not inquire whence [were supplied] to their Mother (whom they call the Enthymesis and impulse of the Æon that went astray) so great an amount of tears, or perspiration, or sadness, or that which produced the remainder of matter.

4. For, to attribute the substance of created things to the power and will of Him who is God of all, is worthy both of credit and acceptance. It is also agreeable [to reason], and there may be well said regarding such a belief, that ``the things which are impossible with men are possible with God.'' \gloss{Luke 18:27} While men, indeed, cannot make anything out of nothing, but only out of matter already existing, yet God is in this point pre-eminently superior to men, that He Himself called into being the substance of His creation, when previously it had no existence. But the assertion that matter was produced from the Enthymesis of an Æon going astray, and that the Æon [referred to] was far separated from her Enthymesis, and that, again, her passion and feeling, apart from herself, became matter— is incredible, infatuated, impossible, and untenable.
\section*{The apostles did not commence to preach the Gospel, or to place anything on record until they were endowed with the gifts and power of the Holy Spirit. They preached one God alone, Maker of heaven and earth.}


1. We have learned from none others the plan of our salvation, than from those through whom the Gospel has come down to us, which they did at one time proclaim in public, and, at a later period, by the will of God, handed down to us in the Scriptures, to be the ground and pillar of our faith.  For it is unlawful to assert that they preached before they possessed ``perfect knowledge,'' as some do even venture to say, boasting themselves as improvers of the apostles. For, after our Lord rose from the dead, [the apostles] were invested with power from on high when the Holy Spirit came down [upon them], were filled from all [His gifts], and had perfect knowledge: they departed to the ends of the earth, preaching the glad tidings of the  things [sent] from God to us, and proclaiming the peace of heaven to men, who indeed do all equally and individually possess the Gospel of God. Matthew also issued a written Gospel among the Hebrews  in their own dialect, while Peter and Paul were preaching at Rome, and laying the foundations of the Church. After their departure, Mark, the disciple and interpreter of Peter, did also hand down to us in writing what had been preached by Peter. Luke also, the companion of Paul, recorded in a book the Gospel preached by him. Afterwards, John, the disciple of the Lord, who also had leaned upon His breast, did himself publish a Gospel during his residence at Ephesus in Asia.

2. These have all declared to us that there is one God, Creator of heaven and earth, announced  by the law and the prophets; and one Christ the Son of God. If any one do not agree to these truths, he despises the companions of the Lord; nay more, he despises Christ Himself the Lord; yea, he despises the Father also, and stands self-condemned, resisting and opposing his own salvation, as is the case with all heretics.
\section*{The Holy Ghost, throughout the Old Testament Scriptures, made mention of no other God or Lord, save him who is the true God.}


1. Therefore neither would the Lord, nor the Holy Spirit, nor the apostles, have ever named as God, definitely and absolutely, him who was not God, unless he were truly God; nor would they have named any one in his own person Lord, except God the Father ruling over all, and His Son who has received dominion from His Father over all creation, as this passage has it: ``The Lord said unto my Lord, Sit at my right hand, until I make Your enemies Your footstool.''  Here the [Scripture] represents to us the Father addressing the Son; He who gave Him the inheritance of the heathen, and subjected to Him all His enemies. Since, therefore, the Father is truly Lord, and the Son truly Lord, the Holy Spirit has fitly designated them by the title of Lord. And again, referring to the destruction of the Sodomites, the Scripture says, ``Then the Lord rained upon Sodom and upon Gomorrha fire and brimstone from the Lord out of heaven.'' \gloss{Genesis 19:24} For it here points out that the Son, who had also been talking with Abraham, had received power to judge the Sodomites  for their wickedness. And this [text following] does declare the same truth: ``Your throne, O God, is for ever and ever; the sceptre of Your kingdom is a right sceptre. You have loved righteousness, and hated iniquity: therefore God, Your God, has anointed You.''  For the Spirit designates both [of them] by the name, of God— both Him who is anointed as Son, and Him who does anoint, that is, the Father. And again: ``God stood in the congregation of the gods, He judges among the gods.''  He [here] refers to the Father and the Son, and those who have received the adoption; but these are the Church. For she is the synagogue of God, which God— that is, the Son Himself— has gathered by Himself. Of whom He again speaks: ``The God of gods, the Lord has spoken, and has called the earth.''  Who is meant by God? He of whom He has said, ``God shall come openly, our God, and shall not keep silence;''  that is, the Son, who came manifested to men who said, ``I have openly appeared to those who seek Me not.'' \gloss{Isaiah 65:1} But of what gods [does he speak]? [Of those] to whom He says, ``I have said, You are gods, and all sons of the Most High.''  To those, no doubt, who have received the grace of the ``adoption, by which we cry, Abba Father.'' \gloss{Romans 8:15} 

2. Wherefore, as I have already stated, no other is named as God, or is called Lord, except Him who is God and Lord of all, who also said to Moses, ``I am that I am. And thus shall you say to the children of Israel: He who is, has sent me unto you;'' \gloss{Exodus 3:14} and His Son Jesus Christ our Lord, who makes those that believe in His name the sons of God. And again, when the Son speaks to Moses, He says, ``I have come down to deliver this people.'' \gloss{Exodus 3:8} For it is He who descended and ascended for the salvation of men. Therefore God has been declared through the Son, who is in the Father, and has the Father in Himself — He who is, the Father bearing witness to the Son, and the Son announcing the Father.— As also Esaias says, ``I too am witness,'' he declares, ``says the Lord God, and the Son whom I have chosen, that you may know, and believe, and understand that I am.'' \gloss{Isaiah 43:10} 

3. When, however, the Scripture terms them [gods] which are no gods, it does not, as I have already remarked, declare them as gods in every sense, but with a certain addition and signification, by which they are shown to be no gods at all. As with David: ``The gods of the heathen are idols of demons;''  and, ``You shall not follow other gods.''  For in that he says ``the gods of the heathen''— but the heathen are ignorant of the true God— and calls them ``other gods,'' he bars their claim [to be looked upon] as gods at all. But as to what they are in their own person, he speaks concerning them; ``for they are,'' he says, ``the idols of demons.'' And Esaias: ``Let them be confounded, all who blaspheme God, and carve useless things;  even I am witness, says God.'' \gloss{Isaiah 44:9} He removes them from [the category of] gods, but he makes use of the word alone, for this [purpose], that we may know of whom he speaks. Jeremiah also says the same: ``The gods that have not made the heavens and earth, let them perish from the earth which is under the heaven.'' \gloss{Jeremiah 10:11} For, from the fact of his having subjoined their destruction, he shows them to be no gods at all. Elias, too, when all Israel was assembled at Mount Carmel, wishing to turn them from idolatry, says to them, ``How long do you halt between two opinions?  If the Lord be God,  follow Him.'' \gloss{1 Kings 18:21, etc.} And again, at the burnt-offering, he thus addresses the idolatrous priests: ``You shall call upon the name of your gods, and I will call on the name of the Lord my God; and the Lord that will hearken by fire,  He is God.'' Now, from the fact of the prophet having said these words, he proves that these gods which were reputed so among those men, are no gods at all. He directed them to that God upon whom he believed, and who was truly God; whom invoking, he exclaimed, ``Lord God of Abraham, God of Isaac, and God of Jacob, hear me today, and let all this people know that You are the God of Israel.'' \gloss{1 Kings 18:36} 

4. Wherefore I do also call upon you, Lord God of Abraham, and God of Isaac, and God of Jacob and Israel, who is the Father of our Lord Jesus Christ, the God who, through the abundance of Your mercy, has  had a favour towards us, that we should know You, who has  made heaven and earth, who rule over all, who is the only and the true God, above whom there is none other God; grant, by our Lord Jesus Christ, the governing power of the Holy Spirit; give to every reader of this book to know You, that You are God alone, to be strengthened in You, and to avoid every heretical, and godless, and impious doctrine.

  5. And the Apostle Paul also, saying, ``For though you have served them which are no gods; you now know God, or rather, are known of God,'' \gloss{Galatians 4:8-9} has made a separation between those that were not [gods] and Him who is God. And again, speaking of Antichrist, he says, ``who opposes and exalts himself above all that is called God, or that is worshipped.'' \gloss{2 Thessalonians 2:4} He points out here those who are called gods, by such as know not God, that is, idols. For the Father of all is called God, and is so; and Antichrist shall be lifted up, not above Him, but above those which are indeed called gods, but are not. And Paul himself says that this is true: ``We know that an idol is nothing, and that there is none other God but one. For though there be that are called gods, whether in heaven or in earth; yet to us there is but one God, the Father, of whom are all things, and we through Him; and one Lord Jesus Christ, by whom are all things, and we by Him.'' \gloss{1 Corinthians 8:4, etc.} For he has made a distinction, and separated those which are indeed called gods, but which are none, from the one God the Father, from whom are all things, and, he has confessed in the most decided manner in his own person, one Lord Jesus Christ. But in this [clause], ``whether in heaven or in earth,'' he does not speak of the formers of the world, as these [teachers] expound it; but his meaning is similar to that of Moses, when it is said, ``You shall not make to yourself any image for God, of whatsoever things are in heaven above, whatsoever in the earth beneath, and whatsoever in the waters under the earth.'' \gloss{Deuteronomy 5:8} And he does thus explain what are meant by the things in heaven: ``Lest when,'' he says, ``looking towards heaven, and observing the sun, and the moon, and the stars, and all the ornament of heaven, falling into error, you should adore and serve them.'' \gloss{Deuteronomy 4:19} And Moses himself, being a man of God, was indeed given as a god before Pharaoh; \gloss{Exodus 7:1} but he is not properly termed Lord, nor is called God by the prophets, but is spoken of by the Spirit as ``Moses, the faithful minister and servant of God,'' \gloss{Hebrews 3:5; Numbers 12:7} which also he was.
\section*{Proofs from the apostolic writings, that Jesus Christ was one and the same, the only begotten Son of God, perfect God and perfect man.}


1. But  there are some who say that Jesus was merely a receptacle of Christ, upon whom the Christ, as a dove, descended from above, and that when He had declared the unnameable Father He entered into the Pleroma in an incomprehensible and invisible manner: for that He was not comprehended, not only by men, but not even by those powers and virtues which are in heaven, and that Jesus was the Son, but that  Christ was the Father, and the Father of Christ, God; while others say that He merely suffered in outward appearance, being naturally impassible. The Valentinians, again, maintain that the dispensational Jesus was the same who passed through Mary, upon whom that Saviour from the more exalted [region] descended, who was also termed \textit{Pan},  because He possessed the names (\textit{vocabula}) of all those who had produced Him; but that [this latter] shared with Him, the dispensational one, His power and His name; so that by His means death was abolished, but the Father was made known by that Saviour who had descended from above, whom they do also allege to be Himself the receptacle of Christ and of the entire Pleroma; confessing, indeed, in tongue one Christ Jesus, but being divided in [actual] opinion: for, as I have already observed, it is the practice of these men to say that there was one Christ, who was produced by Monogenes, for the confirmation of the Pleroma; but that another, the Saviour, was sent [forth] for the glorification of the Father; and yet another, the dispensational one, and whom they represent as having suffered, who also bore [in himself] Christ, that Saviour who returned into the Pleroma. I judge it necessary therefore to take into account the entire mind of the apostles regarding our Lord Jesus Christ, and to show that not only did they never hold any such opinions regarding Him; but, still further, that they announced through the Holy Spirit, that those who should teach such doctrines were agents of Satan, sent forth for the purpose of overturning the faith of some, and drawing them away from life.

2. That John knew the one and the same Word of God, and that He was the only begotten, and that He became incarnate for our salvation, Jesus Christ our Lord, I have sufficiently proved from the word of John himself. And Matthew, too, recognising one and the same Jesus Christ, exhibiting his generation as a man from the Virgin,  even as God did promise David that He would raise up from the fruit of his body an eternal King, having made the same promise to Abraham a long time previously, says: ``The book of the generation of Jesus Christ, the son of David, the son of Abraham.'' \gloss{Matthew 1:1} Then, that he might free our mind from suspicion regarding Joseph, he says: ``But the birth of Christ  was on this  wise. When His mother was espoused to Joseph, before they came together, she was found with child of the Holy Ghost.'' Then, when Joseph had it in contemplation to put Mary away, since she proved with child, [Matthew tells us of] the angel of God standing by him, and saying: ``Fear not to take unto you Mary your wife: for that which is conceived in her is of the Holy Ghost. And she shall bring forth a son, and you shall call His name Jesus; for He shall save His people from their sins. Now this was done, that it might be fulfilled which was spoken of the Lord by the prophet: Behold, a virgin shall conceive, and bring forth a son, and they shall call His name Emmanuel, which is, God with us;'' clearly signifying that both the promise made to the fathers had been accomplished, that the Son of God was born of a virgin, and that He Himself was Christ the Saviour whom the prophets had foretold; not, as these men assert, that Jesus was He who was born of Mary, but that Christ was He who descended from above. Matthew might certainly have said, ``Now the birth of \textit{Jesus} was on this wise;'' but the Holy Ghost, foreseeing the corrupters [of the truth], and guarding by anticipation against their deceit, says by Matthew, ``But the birth of \textit{Christ} was on this wise;'' and that He is Emmanuel, lest perchance we might consider Him as a mere man: for ``not by the will of the flesh nor by the will of man, but by the will of God was the Word made flesh;''  and that we should not imagine that Jesus was one, and Christ another, but should know them to be one and the same.

3. Paul, when writing to the Romans, has explained this very point: ``Paul, an apostle of Jesus Christ, predestinated unto the Gospel of God, which He had promised by His prophets in the holy Scriptures, concerning His Son, who was made to Him of the seed of David according to the flesh, who was predestinated the Son of God with power through the Spirit of holiness, by the resurrection from the dead of our Lord Jesus Christ.'' \gloss{Romans 1:1-4} And again, writing to the Romans about Israel, he says: ``Whose are the fathers, and from whom is Christ according to the flesh, who is God over all, blessed for ever.'' \gloss{Romans 9:5} And again, in his Epistle to the Galatians, he says: ``But when the fullness of time had come, God sent forth His Son, made of a woman, made under the law, to redeem them that were under the law, that we might receive the adoption;'' \gloss{Galatians 4:4-5} plainly indicating one God, who did by the prophets make promise of the Son, and one Jesus Christ our Lord, who was of the seed of David according to His birth from Mary; and that Jesus Christ was appointed the Son of God with power, according to the Spirit of holiness, by the resurrection from the dead, as being the first begotten in all the creation; \gloss{Colossians 1:14-15} the Son of God being made the Son of man, that through Him we may receive the adoption,— humanity  sustaining, and receiving, and embracing the Son of God. Wherefore Mark also says: ``The beginning of the Gospel of Jesus Christ, the Son of God; as it is written in the prophets.'' \gloss{Mark 1:1} Knowing one and the same Son of God, Jesus Christ, who was announced by the prophets, who from the fruit of David's body was Emmanuel, ``the messenger of great counsel of the Father;''  through whom God caused the day-spring and the Just One to arise to the house of David, and raised up for him an horn of salvation, ``and established a testimony in Jacob;'' \gloss{Luke 1:69} as David says when discoursing on the causes of His birth: ``And He appointed a law in Israel, that another generation might know [Him,] the children which should he born from these, and they arising shall themselves declare to their children, so that they might set their hope in God, and seek after His commandments.''  And again, the angel said, when bringing good tidings to Mary: ``He shall he great, and shall be called the Son of the Highest; and the Lord shall give unto Him the throne of His father David;'' \gloss{Luke 1:32} acknowledging that He who is the Son of the Highest, the same is Himself also the Son of David. And David, knowing by the Spirit the dispensation of the advent of this Person, by which He is supreme over all the living and dead, confessed Him as Lord, sitting on the right hand of the Most High Father. 

4. But Simeon also — he who had received an intimation from the Holy Ghost that he should not see death, until first he had beheld Christ Jesus— taking Him, the first-begotten of the Virgin, into his hands, blessed God, and said, ``Lord, now let Your servant depart in peace, according to Your word: because my eyes have seen Your salvation, which You have prepared before the face of all people; a light to lighten the Gentiles, and the glory of Your people Israel;'' \gloss{Luke 2:29} confessing thus, that the infant whom he was holding in his hands, Jesus, born of Mary, was Christ Himself, the Son of God, the light of all, the glory of Israel itself,  and the peace and refreshing of those who had fallen asleep. For He was already despoiling men, by removing their ignorance, conferring upon them His own knowledge, and scattering abroad those who recognised Him, as Esaias says: ``Call His name, Quickly spoil, Rapidly divide.'' \gloss{Isaiah 8:3} Now these are the works of Christ. He therefore was Himself Christ, whom Simeon carrying [in his arms] blessed the Most High; on beholding whom the shepherds glorified God; whom John, while yet in his mother's womb, and He (Christ) in that of Mary, recognising as the Lord, saluted with leaping; whom the Magi, when they had seen, adored, and offered their gifts [to Him], as I have already stated, and prostrated themselves to the eternal King, departed by another way, not now returning by the way of the Assyrians. ``For before the child shall have knowledge to cry, Father or mother, He shall receive the power of Damascus, and the spoils of Samaria, against the king of the Assyrians;'' \gloss{Isaiah 8:4} declaring, in a mysterious manner indeed, but emphatically, that the Lord did fight with a hidden hand against Amalek.  For this cause, too, He suddenly removed those children belonging to the house of David, whose happy lot it was to have been born at that time, that He might send them on before into His kingdom; He, since He was Himself an infant, so arranging it that human infants should be martyrs, slain, according to the Scriptures, for the sake of Christ, who was born in Bethlehem of Judah, in the city of David. \gloss{Matthew 2:16} 

5. Therefore did the Lord also say to His disciples after the resurrection, ``O thoughtless ones, and slow of heart to believe all that the prophets have spoken! Ought not Christ to have suffered these things, and to enter into His glory?'' \gloss{Luke 24:25} And again does He say to them: ``These are the words which I spoke unto you while I was yet with you, that all things must be fulfilled which were written in the law of Moses, and in the prophets, and in the Psalms, concerning Me. Then opened He their understanding, that they should understand the Scriptures, and said unto them, Thus it is written, and thus it behooved Christ to suffer, and to rise again from the dead, and that repentance for the remission of sins be preached in His name among all nations.'' \gloss{Luke 24:44, etc.} Now this is He who was born of Mary; for He says: ``The Son of man must suffer many things, and be rejected, and crucified, and on the third day rise again.''  The Gospel, therefore, knew no other son of man but Him who was of Mary, who also suffered; and no Christ who flew away from Jesus before the passion; but Him who was born it knew as Jesus Christ the Son of God, and that this same suffered and rose again, as John, the disciple of the Lord, verifies, saying: ``But these are written, that you might believe that Jesus is the Christ, the Son of God, and that believing you might have eternal life in His name,'' \gloss{John 20:31} — foreseeing these blasphemous systems which divide the Lord, as far as lies in their power, saying that He was formed of two different substances. For this reason also he has thus testified to us in his Epistle: ``Little children, it is the last time; and as you have heard that Antichrist does come, now have many antichrists appeared; whereby we know that it is the last time. They went out from us, but they were not of us; for if they had been of us, they would have continued with us: but [they departed], that they might be made manifest that they are not of us. Know therefore, that every lie is from without, and is not of the truth. Who is a liar, but he that denies that Jesus is the Christ? This is Antichrist.'' 

6. But inasmuch as all those before mentioned, although they certainly do with their tongue confess one Jesus Christ, make fools of themselves, thinking one thing and saying another;  for their hypotheses vary, as I have already shown, alleging, [as they do,] that one Being suffered and was born, and that this was Jesus; but that there was another who descended upon Him, and that this was Christ, who also ascended again; and they argue, that he who proceeded from the Demiurge, or he who was dispensational, or he who sprang from Joseph, was the Being subject to suffering; but upon the latter there descended from the invisible and ineffable [places] the former, whom they assert to be incomprehensible, invisible, and impassible: they thus wander from the truth, because their doctrine departs from Him who is truly God, being ignorant that His only-begotten Word, who is always present with the human race, united to and mingled with His own creation, according to the Father's pleasure, and who became flesh, is Himself Jesus Christ our Lord, who did also suffer for us, and rose again on our behalf, and who will come again in the glory of His Father, to raise up all flesh, and for the manifestation of salvation, and to apply the rule of just judgment to all who were made by Him. There is therefore, as I have pointed out, one God the Father, and one Christ Jesus, who came by means of the whole dispensational arrangements [connected with Him], and gathered  together all things in Himself. \gloss{Ephesians 1:10} But in every respect, too, He is man, the formation of God; and thus He took up man into Himself, the invisible becoming visible, the incomprehensible being made comprehensible, the impassible becoming capable of suffering, and the Word being made man, thus summing up all things in Himself: so that as in super-celestial, spiritual, and invisible things, the Word of God is supreme, so also in things visible and corporeal He might possess the supremacy, and, taking to Himself the pre-eminence, as well as constituting Himself Head of the Church, He might draw all things to Himself at the proper time.

7. With Him is nothing incomplete or out of due season, just as with the Father there is nothing incongruous. For all these things were foreknown by the Father; but the Son works them out at the proper time in perfect order and sequence. This was the reason why, when Mary was urging [Him] on to [perform] the wonderful miracle of the wine, and was desirous before the time to partake  of the cup of emblematic significance, the Lord, checking her untimely haste, said, ``Woman, what have I to do with you? My hour is not yet come'' \gloss{John 2:4} — waiting for that hour which was foreknown by the Father. This is also the reason why, when men were often desirous to take Him, it is said, ``No man laid hands upon Him, for the hour of His being taken was not yet come;'' \gloss{John 7:30} nor the time of His passion, which had been foreknown by the Father; as also says the prophet Habakkuk, ``By this You shall be known when the years have drawn near; You shall be set forth when the time comes; because my soul is disturbed by anger, You shall remember Your mercy.'' \gloss{Habakkuk 3:2} Paul also says: ``But when the fullness of time came, God sent forth His Son.'' \gloss{Galatians 4:4} By which is made manifest, that all things which had been foreknown of the Father, our Lord did accomplish in their order, season, and hour, foreknown and fitting, being indeed one and the same, but rich and great. For He fulfils the bountiful and comprehensive will of His Father, inasmuch as He is Himself the Saviour of those who are saved, and the Lord of those who are under authority, and the God of all those things which have been formed, the only-begotten of the Father, Christ who was announced, and the Word of God, who became incarnate when the fullness of time had come, at which the Son of God had to become the Son of man.

8. All, therefore, are outside of the [Christian] dispensation, who, under pretext of knowledge, understand that Jesus was one, and Christ another, and the Only-begotten another, from whom again is the Word, and that the Saviour is another, whom these disciples of error allege to be a production of those who were made Æons in a state of degeneracy. Such men are to outward appearance sheep; for they appear to be like us, by what they say in public, repeating the same words as we do; but inwardly they are wolves. Their doctrine is homicidal, conjuring up, as it does, a number of gods, and simulating many Fathers, but lowering and dividing the Son of God in many ways. These are they against whom the Lord has cautioned us beforehand; and His disciple, in his Epistle already mentioned, commands us to avoid them, when he says: ``For many deceivers are entered into the world, who confess not that Jesus Christ has come in the flesh. This is a deceiver and an antichrist. Take heed to them, that you lose not what you have wrought.''  And again does he say in the Epistle: ``Many false prophets are gone out into the world. Hereby know the Spirit of God: Every spirit that confesses that Jesus Christ has come in the flesh is of God; and every spirit which separates Jesus Christ is not of God, but is of antichrist.''  These words agree with what was said in the Gospel, that ``the Word was made flesh, and dwelt among us.'' Wherefore he again exclaims in his Epistle, ``Every one that believes that Jesus is the Christ, has been born of God;'' \gloss{1 John 5:1} knowing Jesus Christ to be one and the same, to whom the gates of heaven were opened, because of His taking upon Him flesh: who shall also come in the same flesh in which He suffered, revealing the glory of the Father.

9. Concurring with these statements, Paul, speaking to the Romans, declares: ``Much more they who receive abundance of grace and righteousness for [eternal] life, shall reign by one, Christ Jesus.'' \gloss{Romans 5:17} It follows from this, that he knew nothing of that Christ who flew away from Jesus; nor did he of the Saviour above, whom they hold to be impassible. For if, in  truth, the one suffered, and the other remained incapable of suffering, and the one was born, but the other descended upon him who was born, and left him again, it is not one, but two, that are shown forth. But that the apostle did know Him as one, both who was born and who suffered, namely Christ Jesus, he again says in the same Epistle: ``Do you not know, that so many of us as were baptized in Christ Jesus were baptized in His death? That like as Christ rose from the dead, so should we also walk in newness of life.'' \gloss{Romans 6:3-4} But again, showing that Christ did suffer, and was Himself the Son of God, who died for us, and redeemed us with His blood at the time appointed beforehand, he says: ``For how is it, that Christ, when we were yet without strength, in due time died for the ungodly? But God commends His love towards us, in that, while we were yet sinners, Christ died for us. Much more, then, being now justified by His blood, we shall be saved from wrath through Him. For if, when we were enemies, we were reconciled to God by the death of His Son; much more, being reconciled, we shall be saved by His life.''  He declares in the plainest manner, that the same Being who was laid hold of, and underwent suffering, and shed His blood for us, was both Christ and the Son of God, who did also rise again, and was taken up into heaven, as he himself [Paul] says: ``But at the same time, [it, is] Christ [that] died, yea rather, that is risen again, who is even at the right hand of God.'' \gloss{Romans 8:34} And again, ``Knowing that Christ, rising from the dead, dies no more:'' \gloss{Romans 6:9} for, as himself foreseeing, through the Spirit, the subdivisions of evil teachers [with regard to the Lord's person], and being desirous of cutting away from them all occasion of cavil, he says what has been already stated, [and also declares:] ``But if the Spirit of Him that raised up Jesus from the dead dwell in you, He that raised up Christ from the dead shall also quicken your mortal bodies.'' \gloss{Romans 8:11} This he does not utter to those alone who wish to hear: Do not err, [he says to all:] Jesus Christ, the Son of God, is one and the same, who did by suffering reconcile us to God, and rose from the dead; who is at the right hand of the Father, and perfect in all things; ``who, when He was buffeted, struck not in return; who, when He suffered, threatened not;'' \gloss{1 Peter 2:23} and when He underwent tyranny, He prayed His Father that He would forgive those who had crucified Him. For He did Himself truly bring in salvation: since He is Himself the Word of God, Himself the Only-begotten of the Father, Christ Jesus our Lord.
\section*{Continuation of the foregoing argument. Proofs from the writings of St. Paul, and from the words of Our Lord, that Christ and Jesus cannot be considered as distinct beings; neither can it be alleged that the Son of God became man merely in appearance, but that He did so truly and actually.}


1.  As it has been clearly demonstrated that  the Word, who existed in the beginning with God, by whom all things were made, who was also always present with mankind, was in these last days, according to the time appointed by the Father, united to His own workmanship, inasmuch as He became a man liable to suffering, [it follows] that every objection is set aside of those who say, ``If our Lord was born at that time, Christ had therefore no previous existence.'' For I have shown that the Son of God did not then begin to exist, being with the Father from the beginning; but when He became incarnate, and was made man, He commenced afresh  the long line of human beings, and furnished us, in a brief, comprehensive manner, with salvation; so that what we had lost in Adam— namely, to be according to the image and likeness of God— that we might recover in Christ Jesus.

2. For as it was not possible that the man who had once for all been conquered, and who had been destroyed through disobedience, could reform himself, and obtain the prize of victory; and as it was also impossible that he could attain to salvation who had fallen under the power of sin—the Son effected both these things, being the Word of God, descending from the Father, becoming incarnate, stooping low, even to death, and consummating the arranged plan of our salvation, upon whom [Paul], exhorting us unhesitatingly to believe, again says, ``Who shall ascend into heaven? That is, to bring down Christ; or who shall descend into the deep? That is, to liberate Christ again from the dead.'' \gloss{Romans 10:6-7} Then he continues, ``If you shall confess with your mouth the Lord Jesus, and shall believe in your heart that God has raised Him from the dead, you shall be saved.'' \gloss{Romans 10:9} And he renders the reason why the Son of God did these things, saying, ``For to this end Christ both lived, and died, and revived, that He might rule over the living and the dead.'' \gloss{Romans 14:9} And again, writing to the Corinthians, he declares, ``But we preach Christ Jesus crucified;'' \gloss{1 Corinthians 1:23} and adds, ``The cup of blessing which we bless, is it not the communion of the blood of Christ?'' \gloss{1 Corinthians 10:16} 

3. But who is it that has had fellowship with us in the matter of food? Whether is it he who is conceived of by them as the Christ above, who extended himself through Horos, and imparted a form to their mother; or is it He who is from the Virgin, Emmanuel, who ate butter and honey, \gloss{Isaiah 8:14} of whom the prophet declared, ``He is also a man, and who shall know him?'' \gloss{Jeremiah 17:9} He was likewise preached by Paul: ``For I delivered,'' he says, ``unto you first of all, that Christ died for our sins, according to the Scriptures; and that He was buried, and rose again the third day, according to the Scriptures.'' \gloss{1 Corinthians 15:3-4} It is plain, then, that Paul knew no other Christ besides Him alone, who both suffered, and was buried, and rose gain, who was also born, and whom he speaks of as man. For after remarking, ``But if Christ be preached, that He rose from the dead,'' \gloss{1 Corinthians 15:12} he continues, rendering the reason of His incarnation, ``For since by man came death, by man [came] also the resurrection of the dead.'' And everywhere, when [referring to] the passion of our Lord, and to His human nature, and His subjection to death, he employs the name of Christ, as in that passage: ``Destroy not him with your meat for whom Christ died.'' \gloss{Romans 14:15} And again: ``But now, in Christ, you who sometimes were far off are made near by the blood of Christ.'' \gloss{Ephesians 2:13} And again: ``Christ has redeemed us from the curse of the law, being made a curse for us: for it is written, Cursed is every one that hangs upon a tree.'' \gloss{Galatians 3:13; Deuteronomy 21:23} And again: ``And through your knowledge shall the weak brother perish, for whom Christ died;'' \gloss{1 Corinthians 8:11} indicating that the impassible Christ did not descend upon Jesus, but that He Himself, because He was Jesus Christ, suffered for us; He, who lay in the tomb, and rose again, who descended and ascended,— the Son of God having been made the Son of man, as the very name itself does declare. For in the name of Christ is implied, He that anoints, He that is anointed, and the unction itself with which He is anointed. And it is the Father who anoints, but the Son who is anointed by the Spirit, who is the unction, as the Word declares by Isaiah, ``The Spirit of the Lord is upon me, because He has anointed me,'' \gloss{Isaiah 61:1} — pointing out both the anointing Father, the anointed Son, and the unction, which is the Spirit.

4. The Lord Himself, too, makes it evident who it was that suffered; for when He asked the disciples, ``Who do men say that I, the Son of man, am?'' \gloss{Matthew 16:13} and when Peter had replied, ``You are the Christ, the Son of the living God;'' and when he had been commended by Him [in these words], ``That flesh and blood had not revealed it to him, but the Father who is in heaven,'' He made it clear that He, the Son of man, is Christ the Son of the living God. ``For from that time forth,'' it is said, ``He began to show to His disciples, how that He must go unto Jerusalem,  and suffer many things of the priests, and be rejected, and crucified, and rise again the third day.'' \gloss{Matthew 16:21} He who was acknowledged by Peter as Christ, who pronounced him blessed because the Father had revealed the Son of the living God to him, said that He must Himself suffer many things, and be crucified; and then He rebuked Peter, who imagined that He was the Christ as the generality of men supposed  [that the Christ should be], and was averse to the idea of His suffering, [and] said to the disciples, ``If any man will come after Me, let him deny himself, and take up his cross, and follow Me. For whosoever will save his life, shall lose it; and whosoever will lose it for My sake shall save it.'' \gloss{Matthew 16:24-25} For these things Christ spoke openly, He being Himself the Saviour of those who should be delivered over to death for their confession of Him, and lose their lives.

5. If, however, He was Himself not to suffer, but should fly away from Jesus, why did He exhort His disciples to take up the cross and follow Him—that cross which these men represent Him as not having taken up, but [speak of Him] as having relinquished the dispensation of suffering? For that He did not say this with reference to the acknowledging of the \textit{Stauros} (cross) above, as some among them venture to expound, but with respect to the suffering which He should Himself undergo, and that His disciples should endure, He implies when He says, ``For whosoever will save his life, shall lose it; and whosoever will lose, shall find it.'' And that His disciples must suffer for His sake, He [implied when He] said to the Jews, ``Behold, I send you prophets, and wise men, and scribes: and some of them you shall kill and crucify.'' \gloss{Matthew 23:24} And to the disciples He was wont to say, ``And you shall stand before governors and kings for My sake; and they shall scourge some of you, and slay you, and persecute you from city to city.'' \gloss{Matthew 10:17-18} He knew, therefore, both those who should suffer persecution, and He knew those who should have to be scourged and slain because of Him; and He did not speak of any other cross, but of the suffering which He should Himself undergo first, and His disciples afterwards. For this purpose did He give them this exhortation: ``Fear not them which kill the body, but are not able to kill the soul; but rather fear Him who is able to send both soul and body into hell;'' \gloss{Matthew 10:28} [thus exhorting them] to hold fast those professions of faith which they had made in reference to Him. For He promised to confess before His Father those who should confess His name before men; but declared that He would deny those who should deny Him, and would be ashamed of those who should be ashamed to confess Him. And although these things are so, some of these men have proceeded to such a degree of temerity, that they even pour contempt upon the martyrs, and vituperate those who are slain on account of the confession of the Lord, and who suffer all things predicted by the Lord, and who in this respect strive to follow the footprints of the Lord's passion, having become martyrs of the suffering One; these we do also enrol with the martyrs themselves. For, when inquisition shall be made for their blood,  and they shall attain to glory, then all shall be confounded by Christ, who have cast a slur upon their martyrdom. And from this fact, that He exclaimed upon the cross, ``Father, forgive them, for they know not what they do,'' \gloss{Luke 23:34} the long-suffering, patience, compassion, and goodness of Christ are exhibited, since He both suffered, and did Himself exculpate those who had maltreated Him. For the Word of God, who said to us, ``Love your enemies, and pray for those that hate you,'' \gloss{Matthew 5:44} Himself did this very thing upon the cross; loving the human race to such a degree, that He even prayed for those putting Him to death. If, however, any one, going upon the supposition that there are two [Christs], forms a judgment in regard to them, that [Christ] shall be found much the better one, and more patient, and the truly good one, who, in the midst of His own wounds and stripes, and the other [cruelties] inflicted upon Him, was beneficent, and unmindful of the wrongs perpetrated upon Him, than he who flew away, and sustained neither injury nor insult.

6. This also does likewise meet [the case] of those who maintain that He suffered only in appearance. For if He did not truly suffer, no thanks to Him, since there was no suffering at all; and when we shall actually begin to suffer, He will seem as leading us astray, exhorting us to endure buffering, and to turn the other \gloss{Matthew 5:39} cheek, if He did not Himself before us in reality suffer the same; and as He misled them by seeming to them what He was not, so does He also mislead us, by exhorting us to endure what He did not endure Himself. [In that case] we shall be even above the Master, because we suffer and sustain what our Master never bore or endured. But as our Lord is alone truly Master, so the Son of God is truly good and patient, the Word of God the Father having been made the Son of man. For He fought and conquered; for He was man contending for the fathers,  and  through obedience doing away with disobedience completely: for He bound the strong man, \gloss{Matthew 12:29} and set free the weak, and endowed His own handiwork with salvation, by destroying sin. For He is a most holy and merciful Lord, and loves the human race.

7. Therefore, as I have already said, He caused man (human nature) to cleave to and to become, one with God. For unless man had overcome the enemy of man, the enemy would not have been legitimately vanquished. And again: unless it had been God who had freely given salvation, we could never have possessed it securely. And unless man had been joined to God, he could never have become a partaker of incorruptibility. For it was incumbent upon the Mediator between God and men, by His relationship to both, to bring both to friendship and concord, and present man to God, while He revealed God to man.  For, in what way could we be partaken of the adoption of sons, unless we had received from Him through the Son that fellowship which refers to Himself, unless His Word, having been made flesh, had entered into communion with us? Wherefore also He passed through every stage of life, restoring to all communion with God. Those, therefore, who assert that He appeared putatively, and was neither born in the flesh nor truly made man, are as yet under the old condemnation, holding out patronage to sin; for, by their showing, death has not been vanquished, which ``reigned from Adam to Moses, even over them that had not sinned after the similitude of Adam's transgression.'' \gloss{Romans 5:14} But the law coming, which was given by Moses, and testifying of sin that it is a sinner, did truly take away his (death's) kingdom, showing that he was no king, but a robber; and it revealed him as a murderer. It laid, however, a weighty burden upon man, who had sin in himself, showing that he was liable to death. For as the law was spiritual, it merely made sin to stand out in relief, but did not destroy it. For sin had no dominion over the spirit, but over man. For it behooved Him who was to destroy sin, and redeem man under the power of death, that He should Himself be made that very same thing which he was, that is, man; who had been drawn by sin into bondage, but was held by death, so that sin should be destroyed by man, and man should go forth from death. For as by the disobedience of the one man who was originally moulded from virgin soil, the many were made sinners, \gloss{Romans 5:19} and forfeited life; so was it necessary that, by the obedience of one man, who was originally born from a virgin, many should be justified and receive salvation. Thus, then, was the Word of God made man, as also Moses says: ``God, true are His works.'' \gloss{Deuteronomy 32:4} But if, not having been made flesh, He did appear as if flesh, His work was not a true one. But what He did appear, that He also was: God recapitulated in Himself the ancient formation of man, that He might kill sin, deprive death of its power, and vivify man; and therefore His works are true.
\section*{Proofs from the apostolic writings, that Jesus Christ was one and the same, the only begotten Son of God, perfect God and perfect man.}


1. But  there are some who say that Jesus was merely a receptacle of Christ, upon whom the Christ, as a dove, descended from above, and that when He had declared the unnameable Father He entered into the Pleroma in an incomprehensible and invisible manner: for that He was not comprehended, not only by men, but not even by those powers and virtues which are in heaven, and that Jesus was the Son, but that  Christ was the Father, and the Father of Christ, God; while others say that He merely suffered in outward appearance, being naturally impassible. The Valentinians, again, maintain that the dispensational Jesus was the same who passed through Mary, upon whom that Saviour from the more exalted [region] descended, who was also termed \textit{Pan},  because He possessed the names (\textit{vocabula}) of all those who had produced Him; but that [this latter] shared with Him, the dispensational one, His power and His name; so that by His means death was abolished, but the Father was made known by that Saviour who had descended from above, whom they do also allege to be Himself the receptacle of Christ and of the entire Pleroma; confessing, indeed, in tongue one Christ Jesus, but being divided in [actual] opinion: for, as I have already observed, it is the practice of these men to say that there was one Christ, who was produced by Monogenes, for the confirmation of the Pleroma; but that another, the Saviour, was sent [forth] for the glorification of the Father; and yet another, the dispensational one, and whom they represent as having suffered, who also bore [in himself] Christ, that Saviour who returned into the Pleroma. I judge it necessary therefore to take into account the entire mind of the apostles regarding our Lord Jesus Christ, and to show that not only did they never hold any such opinions regarding Him; but, still further, that they announced through the Holy Spirit, that those who should teach such doctrines were agents of Satan, sent forth for the purpose of overturning the faith of some, and drawing them away from life.

2. That John knew the one and the same Word of God, and that He was the only begotten, and that He became incarnate for our salvation, Jesus Christ our Lord, I have sufficiently proved from the word of John himself. And Matthew, too, recognising one and the same Jesus Christ, exhibiting his generation as a man from the Virgin,  even as God did promise David that He would raise up from the fruit of his body an eternal King, having made the same promise to Abraham a long time previously, says: ``The book of the generation of Jesus Christ, the son of David, the son of Abraham.'' \gloss{Matthew 1:1} Then, that he might free our mind from suspicion regarding Joseph, he says: ``But the birth of Christ  was on this  wise. When His mother was espoused to Joseph, before they came together, she was found with child of the Holy Ghost.'' Then, when Joseph had it in contemplation to put Mary away, since she proved with child, [Matthew tells us of] the angel of God standing by him, and saying: ``Fear not to take unto you Mary your wife: for that which is conceived in her is of the Holy Ghost. And she shall bring forth a son, and you shall call His name Jesus; for He shall save His people from their sins. Now this was done, that it might be fulfilled which was spoken of the Lord by the prophet: Behold, a virgin shall conceive, and bring forth a son, and they shall call His name Emmanuel, which is, God with us;'' clearly signifying that both the promise made to the fathers had been accomplished, that the Son of God was born of a virgin, and that He Himself was Christ the Saviour whom the prophets had foretold; not, as these men assert, that Jesus was He who was born of Mary, but that Christ was He who descended from above. Matthew might certainly have said, ``Now the birth of \textit{Jesus} was on this wise;'' but the Holy Ghost, foreseeing the corrupters [of the truth], and guarding by anticipation against their deceit, says by Matthew, ``But the birth of \textit{Christ} was on this wise;'' and that He is Emmanuel, lest perchance we might consider Him as a mere man: for ``not by the will of the flesh nor by the will of man, but by the will of God was the Word made flesh;''  and that we should not imagine that Jesus was one, and Christ another, but should know them to be one and the same.

3. Paul, when writing to the Romans, has explained this very point: ``Paul, an apostle of Jesus Christ, predestinated unto the Gospel of God, which He had promised by His prophets in the holy Scriptures, concerning His Son, who was made to Him of the seed of David according to the flesh, who was predestinated the Son of God with power through the Spirit of holiness, by the resurrection from the dead of our Lord Jesus Christ.'' \gloss{Romans 1:1-4} And again, writing to the Romans about Israel, he says: ``Whose are the fathers, and from whom is Christ according to the flesh, who is God over all, blessed for ever.'' \gloss{Romans 9:5} And again, in his Epistle to the Galatians, he says: ``But when the fullness of time had come, God sent forth His Son, made of a woman, made under the law, to redeem them that were under the law, that we might receive the adoption;'' \gloss{Galatians 4:4-5} plainly indicating one God, who did by the prophets make promise of the Son, and one Jesus Christ our Lord, who was of the seed of David according to His birth from Mary; and that Jesus Christ was appointed the Son of God with power, according to the Spirit of holiness, by the resurrection from the dead, as being the first begotten in all the creation; \gloss{Colossians 1:14-15} the Son of God being made the Son of man, that through Him we may receive the adoption,— humanity  sustaining, and receiving, and embracing the Son of God. Wherefore Mark also says: ``The beginning of the Gospel of Jesus Christ, the Son of God; as it is written in the prophets.'' \gloss{Mark 1:1} Knowing one and the same Son of God, Jesus Christ, who was announced by the prophets, who from the fruit of David's body was Emmanuel, ``the messenger of great counsel of the Father;''  through whom God caused the day-spring and the Just One to arise to the house of David, and raised up for him an horn of salvation, ``and established a testimony in Jacob;'' \gloss{Luke 1:69} as David says when discoursing on the causes of His birth: ``And He appointed a law in Israel, that another generation might know [Him,] the children which should he born from these, and they arising shall themselves declare to their children, so that they might set their hope in God, and seek after His commandments.''  And again, the angel said, when bringing good tidings to Mary: ``He shall he great, and shall be called the Son of the Highest; and the Lord shall give unto Him the throne of His father David;'' \gloss{Luke 1:32} acknowledging that He who is the Son of the Highest, the same is Himself also the Son of David. And David, knowing by the Spirit the dispensation of the advent of this Person, by which He is supreme over all the living and dead, confessed Him as Lord, sitting on the right hand of the Most High Father. 

4. But Simeon also — he who had received an intimation from the Holy Ghost that he should not see death, until first he had beheld Christ Jesus— taking Him, the first-begotten of the Virgin, into his hands, blessed God, and said, ``Lord, now let Your servant depart in peace, according to Your word: because my eyes have seen Your salvation, which You have prepared before the face of all people; a light to lighten the Gentiles, and the glory of Your people Israel;'' \gloss{Luke 2:29} confessing thus, that the infant whom he was holding in his hands, Jesus, born of Mary, was Christ Himself, the Son of God, the light of all, the glory of Israel itself,  and the peace and refreshing of those who had fallen asleep. For He was already despoiling men, by removing their ignorance, conferring upon them His own knowledge, and scattering abroad those who recognised Him, as Esaias says: ``Call His name, Quickly spoil, Rapidly divide.'' \gloss{Isaiah 8:3} Now these are the works of Christ. He therefore was Himself Christ, whom Simeon carrying [in his arms] blessed the Most High; on beholding whom the shepherds glorified God; whom John, while yet in his mother's womb, and He (Christ) in that of Mary, recognising as the Lord, saluted with leaping; whom the Magi, when they had seen, adored, and offered their gifts [to Him], as I have already stated, and prostrated themselves to the eternal King, departed by another way, not now returning by the way of the Assyrians. ``For before the child shall have knowledge to cry, Father or mother, He shall receive the power of Damascus, and the spoils of Samaria, against the king of the Assyrians;'' \gloss{Isaiah 8:4} declaring, in a mysterious manner indeed, but emphatically, that the Lord did fight with a hidden hand against Amalek.  For this cause, too, He suddenly removed those children belonging to the house of David, whose happy lot it was to have been born at that time, that He might send them on before into His kingdom; He, since He was Himself an infant, so arranging it that human infants should be martyrs, slain, according to the Scriptures, for the sake of Christ, who was born in Bethlehem of Judah, in the city of David. \gloss{Matthew 2:16} 

5. Therefore did the Lord also say to His disciples after the resurrection, ``O thoughtless ones, and slow of heart to believe all that the prophets have spoken! Ought not Christ to have suffered these things, and to enter into His glory?'' \gloss{Luke 24:25} And again does He say to them: ``These are the words which I spoke unto you while I was yet with you, that all things must be fulfilled which were written in the law of Moses, and in the prophets, and in the Psalms, concerning Me. Then opened He their understanding, that they should understand the Scriptures, and said unto them, Thus it is written, and thus it behooved Christ to suffer, and to rise again from the dead, and that repentance for the remission of sins be preached in His name among all nations.'' \gloss{Luke 24:44, etc.} Now this is He who was born of Mary; for He says: ``The Son of man must suffer many things, and be rejected, and crucified, and on the third day rise again.''  The Gospel, therefore, knew no other son of man but Him who was of Mary, who also suffered; and no Christ who flew away from Jesus before the passion; but Him who was born it knew as Jesus Christ the Son of God, and that this same suffered and rose again, as John, the disciple of the Lord, verifies, saying: ``But these are written, that you might believe that Jesus is the Christ, the Son of God, and that believing you might have eternal life in His name,'' \gloss{John 20:31} — foreseeing these blasphemous systems which divide the Lord, as far as lies in their power, saying that He was formed of two different substances. For this reason also he has thus testified to us in his Epistle: ``Little children, it is the last time; and as you have heard that Antichrist does come, now have many antichrists appeared; whereby we know that it is the last time. They went out from us, but they were not of us; for if they had been of us, they would have continued with us: but [they departed], that they might be made manifest that they are not of us. Know therefore, that every lie is from without, and is not of the truth. Who is a liar, but he that denies that Jesus is the Christ? This is Antichrist.'' 

6. But inasmuch as all those before mentioned, although they certainly do with their tongue confess one Jesus Christ, make fools of themselves, thinking one thing and saying another;  for their hypotheses vary, as I have already shown, alleging, [as they do,] that one Being suffered and was born, and that this was Jesus; but that there was another who descended upon Him, and that this was Christ, who also ascended again; and they argue, that he who proceeded from the Demiurge, or he who was dispensational, or he who sprang from Joseph, was the Being subject to suffering; but upon the latter there descended from the invisible and ineffable [places] the former, whom they assert to be incomprehensible, invisible, and impassible: they thus wander from the truth, because their doctrine departs from Him who is truly God, being ignorant that His only-begotten Word, who is always present with the human race, united to and mingled with His own creation, according to the Father's pleasure, and who became flesh, is Himself Jesus Christ our Lord, who did also suffer for us, and rose again on our behalf, and who will come again in the glory of His Father, to raise up all flesh, and for the manifestation of salvation, and to apply the rule of just judgment to all who were made by Him. There is therefore, as I have pointed out, one God the Father, and one Christ Jesus, who came by means of the whole dispensational arrangements [connected with Him], and gathered  together all things in Himself. \gloss{Ephesians 1:10} But in every respect, too, He is man, the formation of God; and thus He took up man into Himself, the invisible becoming visible, the incomprehensible being made comprehensible, the impassible becoming capable of suffering, and the Word being made man, thus summing up all things in Himself: so that as in super-celestial, spiritual, and invisible things, the Word of God is supreme, so also in things visible and corporeal He might possess the supremacy, and, taking to Himself the pre-eminence, as well as constituting Himself Head of the Church, He might draw all things to Himself at the proper time.

7. With Him is nothing incomplete or out of due season, just as with the Father there is nothing incongruous. For all these things were foreknown by the Father; but the Son works them out at the proper time in perfect order and sequence. This was the reason why, when Mary was urging [Him] on to [perform] the wonderful miracle of the wine, and was desirous before the time to partake  of the cup of emblematic significance, the Lord, checking her untimely haste, said, ``Woman, what have I to do with you? My hour is not yet come'' \gloss{John 2:4} — waiting for that hour which was foreknown by the Father. This is also the reason why, when men were often desirous to take Him, it is said, ``No man laid hands upon Him, for the hour of His being taken was not yet come;'' \gloss{John 7:30} nor the time of His passion, which had been foreknown by the Father; as also says the prophet Habakkuk, ``By this You shall be known when the years have drawn near; You shall be set forth when the time comes; because my soul is disturbed by anger, You shall remember Your mercy.'' \gloss{Habakkuk 3:2} Paul also says: ``But when the fullness of time came, God sent forth His Son.'' \gloss{Galatians 4:4} By which is made manifest, that all things which had been foreknown of the Father, our Lord did accomplish in their order, season, and hour, foreknown and fitting, being indeed one and the same, but rich and great. For He fulfils the bountiful and comprehensive will of His Father, inasmuch as He is Himself the Saviour of those who are saved, and the Lord of those who are under authority, and the God of all those things which have been formed, the only-begotten of the Father, Christ who was announced, and the Word of God, who became incarnate when the fullness of time had come, at which the Son of God had to become the Son of man.

8. All, therefore, are outside of the [Christian] dispensation, who, under pretext of knowledge, understand that Jesus was one, and Christ another, and the Only-begotten another, from whom again is the Word, and that the Saviour is another, whom these disciples of error allege to be a production of those who were made Æons in a state of degeneracy. Such men are to outward appearance sheep; for they appear to be like us, by what they say in public, repeating the same words as we do; but inwardly they are wolves. Their doctrine is homicidal, conjuring up, as it does, a number of gods, and simulating many Fathers, but lowering and dividing the Son of God in many ways. These are they against whom the Lord has cautioned us beforehand; and His disciple, in his Epistle already mentioned, commands us to avoid them, when he says: ``For many deceivers are entered into the world, who confess not that Jesus Christ has come in the flesh. This is a deceiver and an antichrist. Take heed to them, that you lose not what you have wrought.''  And again does he say in the Epistle: ``Many false prophets are gone out into the world. Hereby know the Spirit of God: Every spirit that confesses that Jesus Christ has come in the flesh is of God; and every spirit which separates Jesus Christ is not of God, but is of antichrist.''  These words agree with what was said in the Gospel, that ``the Word was made flesh, and dwelt among us.'' Wherefore he again exclaims in his Epistle, ``Every one that believes that Jesus is the Christ, has been born of God;'' \gloss{1 John 5:1} knowing Jesus Christ to be one and the same, to whom the gates of heaven were opened, because of His taking upon Him flesh: who shall also come in the same flesh in which He suffered, revealing the glory of the Father.

9. Concurring with these statements, Paul, speaking to the Romans, declares: ``Much more they who receive abundance of grace and righteousness for [eternal] life, shall reign by one, Christ Jesus.'' \gloss{Romans 5:17} It follows from this, that he knew nothing of that Christ who flew away from Jesus; nor did he of the Saviour above, whom they hold to be impassible. For if, in  truth, the one suffered, and the other remained incapable of suffering, and the one was born, but the other descended upon him who was born, and left him again, it is not one, but two, that are shown forth. But that the apostle did know Him as one, both who was born and who suffered, namely Christ Jesus, he again says in the same Epistle: ``Do you not know, that so many of us as were baptized in Christ Jesus were baptized in His death? That like as Christ rose from the dead, so should we also walk in newness of life.'' \gloss{Romans 6:3-4} But again, showing that Christ did suffer, and was Himself the Son of God, who died for us, and redeemed us with His blood at the time appointed beforehand, he says: ``For how is it, that Christ, when we were yet without strength, in due time died for the ungodly? But God commends His love towards us, in that, while we were yet sinners, Christ died for us. Much more, then, being now justified by His blood, we shall be saved from wrath through Him. For if, when we were enemies, we were reconciled to God by the death of His Son; much more, being reconciled, we shall be saved by His life.''  He declares in the plainest manner, that the same Being who was laid hold of, and underwent suffering, and shed His blood for us, was both Christ and the Son of God, who did also rise again, and was taken up into heaven, as he himself [Paul] says: ``But at the same time, [it, is] Christ [that] died, yea rather, that is risen again, who is even at the right hand of God.'' \gloss{Romans 8:34} And again, ``Knowing that Christ, rising from the dead, dies no more:'' \gloss{Romans 6:9} for, as himself foreseeing, through the Spirit, the subdivisions of evil teachers [with regard to the Lord's person], and being desirous of cutting away from them all occasion of cavil, he says what has been already stated, [and also declares:] ``But if the Spirit of Him that raised up Jesus from the dead dwell in you, He that raised up Christ from the dead shall also quicken your mortal bodies.'' \gloss{Romans 8:11} This he does not utter to those alone who wish to hear: Do not err, [he says to all:] Jesus Christ, the Son of God, is one and the same, who did by suffering reconcile us to God, and rose from the dead; who is at the right hand of the Father, and perfect in all things; ``who, when He was buffeted, struck not in return; who, when He suffered, threatened not;'' \gloss{1 Peter 2:23} and when He underwent tyranny, He prayed His Father that He would forgive those who had crucified Him. For He did Himself truly bring in salvation: since He is Himself the Word of God, Himself the Only-begotten of the Father, Christ Jesus our Lord.
\section*{That one God formed all things in the world, by means of the Word and the Holy Spirit: and that although He is to us in this life invisible and incomprehensible, nevertheless He is not unknown; inasmuch as His works do declare Him, and His Word has shown that in many modes He may be seen and known.}


1. As regards His greatness, therefore, it is not possible to know God, for it is impossible that the Father can be measured; but as regards His love (for this it is which leads us to God by His Word), when we obey Him, we do always learn that there is so great a God, and that it is He who by Himself has established, and selected, and adorned, and contains all things; and among the all things, both ourselves and this our world. We also then were made, along with those things which are contained by Him. And this is He of whom the Scripture says, ``And God formed man, taking clay of the earth, and breathed into his face the breath of life.'' \gloss{Genesis 2:7} It was not angels, therefore, who made us, nor who formed us, neither had angels power to make an image of God, nor any one else, except the Word of the Lord, nor any Power remotely distant from the Father of all things. For God did not stand in need of these [beings], in order to the accomplishing of what He had Himself determined with Himself beforehand should be done, as if He did not possess His own hands. For with Him were always present the Word and Wisdom, the Son and the Spirit, by whom and in whom,  freely and spontaneously, He made all things, to whom also He speaks, saying, ``Let Us make man after Our image and likeness;'' \gloss{Genesis 1:26} He taking from Himself the substance of the creatures [formed], and the pattern of things made, and the type of all the adornments in the world.

2. Truly, then, the Scripture declared, which says, ``First  of all believe that there is one God, who has established all things, and completed them, and having caused that from what had no being, all things should come into existence:'' He who contains all things, and is Himself contained by no one. Rightly also has Malachi said among the prophets: ``Is it not one God who has established us? Have we not all one Father?'' \gloss{Malachi 2:10} In accordance with this, too, does the apostle say, ``There is one God, the Father, who is above all, and in us all.'' \gloss{Ephesians 4:6} Likewise does the Lord also say: ``All things are delivered to Me by My Father;'' \gloss{Matthew 11:27} manifestly by Him who made all things; for He did not deliver to Him the things of another, but His own. But in \textit{all things} [it is implied that] nothing has been kept back [from Him], and for this reason the same person is the Judge of the living and the dead; ``having the key of David: He shall open, and no man shall shut: He shall shut, and no man shall open.'' \gloss{Revelation 3:7} For no one was able, either in heaven or in earth, or under the earth, to open the book of the Father, or to behold Him, with the exception of the Lamb who was slain, and who redeemed us with His own blood, receiving power over all things from the same God who made all things by the Word, and adorned them by [His] Wisdom, when ``the Word was made flesh;'' that even as the Word of God had the sovereignty in the heavens, so also might He have the sovereignty in earth, inasmuch as [He was] a righteous man, ``who did no sin, neither was there found guile in His mouth;'' \gloss{1 Peter 2:23} and that He might have the pre-eminence over those things which are under the earth, He Himself being made ``the first-begotten of the dead;'' \gloss{Colossians 1:18} and that all things, as I have already said, might behold their King; and that the paternal light might meet with and rest upon the flesh of our Lord, and come to us from His resplendent flesh, and that thus man might attain to immortality, having been invested with the paternal light.

3. I have also largely demonstrated, that the Word, namely the Son, was always with the Father; and that Wisdom also, which is the Spirit, was present with Him, anterior to all creation, He declares by Solomon: ``God by Wisdom founded the earth, and by understanding has He established the heaven. By His knowledge the depths burst forth, and the clouds dropped down the dew.'' \gloss{Proverbs 3:19-20} And again: ``The Lord created me the beginning of His ways in His work: He set me up from everlasting, in the beginning, before He made the earth, before He established the depths, and before the fountains of waters gushed forth; before the mountains were made strong, and before all the hills, He brought me forth.''  And again: ``When He prepared the heaven, I was with Him, and when He established the fountains of the deep; when He made the foundations of the earth strong, I was with Him preparing [them]. I was He in whom He rejoiced, and throughout all time I was daily glad before His face, when He rejoiced at the completion of the world, and was delighted in the sons of men.'' \gloss{Proverbs 8:27-31} 

4. There is therefore one God, who by the Word and Wisdom created and arranged all things; but this is the Creator (Demiurge) who has granted this world to the human race, and who, as regards His greatness, is indeed unknown to all who have been made by Him (for no man has searched out His height, either among the ancients who have gone to their rest, or any of those who are now alive); but as regards His love, He is always known through Him by whose means He ordained all things. Now this is His Word, our Lord Jesus Christ, who in the last times was made a man among men, that He might join the end to the beginning, that is, man to God. Wherefore the prophets, receiving the prophetic gift from the same Word, announced His advent according to the flesh, by which the blending and communion of God and man took place according to the good pleasure of the Father, the Word of God foretelling from the beginning that God should be seen by men, and hold converse with them upon earth, should confer with them, and should be present with His own creation, saving it, and becoming capable of being perceived by it, and freeing us from the hands of all that hate us, that is, from every spirit of wickedness; and causing us to serve Him in holiness and righteousness all our days, \gloss{Luke 1:71, 75} in order that man, having embraced the Spirit of God, might pass into the glory of the Father.

5. These things did the prophets set forth in a prophetical manner; but they did not, as some allege, [proclaim] that He who was seen by the prophets was a different [God], the Father of  all being invisible. Yet this is what those [heretics] declare, who are altogether ignorant of the nature of prophecy. For prophecy is a prediction of things future, that is, a setting forth beforehand of those things which shall be afterwards. The prophets, then, indicated beforehand that God should be seen by men; as the Lord also says, ``Blessed are the pure in heart, for they shall see God.'' \gloss{Matthew 5:8} But in respect to His greatness, and His wonderful glory, ``no man shall see God and live,'' \gloss{Exodus 33:20} for the Father is incomprehensible; but in regard to His love, and kindness, and as to His infinite power, even this He grants to those who love Him, that is, to see God, which thing the prophets did also predict. ``For those things that are impossible with men, are possible with God.'' \gloss{Luke 18:27} For man does not see God by his own powers; but when He pleases He is seen by men, by whom He wills, and when He wills, and as He wills. For God is powerful in all things, having been seen at that time indeed, prophetically through the Spirit, and seen, too, adoptively through the Son; and He shall also be seen paternally in the kingdom of heaven, the Spirit truly preparing man in the Son  of God, and the Son leading him to the Father, while the Father, too, confers [upon him] incorruption for eternal life, which comes to every one from the fact of his seeing God. For as those who see the light are within the light, and partake of its brilliancy; even so, those who see God are in God, and receive of His splendour. But [His] splendour vivifies them; those, therefore, who see God, do receive life. And for this reason, He, [although] beyond comprehension, and boundless and invisible, rendered Himself visible, and comprehensible, and within the capacity of those who believe, that He might vivify those who receive and behold Him through faith.  For as His greatness is past finding out, so also His goodness is beyond expression; by which having been seen, He bestows life upon those who see Him. It is not possible to live apart from life, and the means of life is found in fellowship with God; but fellowship with God is to know God, and to enjoy His goodness.

6. Men therefore shall see God, that they may live, being made immortal by that sight, and attaining even unto God; which, as I have already said, was declared figuratively by the prophets, that God should be seen by men who bear His Spirit [in them], and do always wait patiently for His coming. As also Moses says in Deuteronomy, ``We shall see in that day that God will talk to man, and he shall live.'' \gloss{Deuteronomy 5:24} For certain of these men used to see the prophetic Spirit and His active influences poured forth for all kinds of gifts; others, again, [beheld] the advent of the Lord, and that dispensation which obtained from the beginning, by which He accomplished the will of the Father with regard to things both celestial and terrestrial; and others [beheld] paternal glories adapted to the times, and to those who saw and who heard them then, and to all who were subsequently to hear them. Thus, therefore, was God revealed; for God the Father is shown forth through all these [operations], the Spirit indeed working, and the Son ministering, while the Father was approving, and man's salvation being accomplished. As He also declares through Hosea the prophet: ``I,'' He says, ``have multiplied visions, and have used similitudes by the ministry (\textit{in manibus}) of the prophets.'' \gloss{Hosea 12:10} But the apostle expounded this very passage, when he said, ``Now there are diversities of gifts, but the same Spirit; and there are differences of ministrations, but the same Lord; and there are diversities of operations, but it is the same God which works all in all. But the manifestation of the Spirit is given to every man to profit withal.'' \gloss{1 Corinthians 12:4-7} But as He who works all things in all is God, [as to the points] of what nature and how great He is, [God] is invisible and indescribable to all things which have been made by Him, but He is by no means unknown: for all things learn through His Word that there is one God the Father, who contains all things, and who grants existence to all, as is written in the Gospel: ``No man has seen God at any time, except the only-begotten Son, who is in the bosom of the Father; He has declared [Him].'' \gloss{John 1:18} 

7. Therefore the Son of the Father declares [Him] from the beginning, inasmuch as He was with the Father from the beginning, who did also show to the human race prophetic visions, and diversities of gifts, and His own ministrations, and the glory of the Father, in regular order and connection, at the fitting time for the benefit [of mankind]. For where there is a regular succession, there is also fixedness; and where fixedness, there suitability to the period; and where suitability, there also utility. And for this reason did the Word become the dispenser of the paternal grace for the benefit of men, for whom He made such great dispensations, revealing God indeed to men, but presenting man to God, and preserving at the same time the invisibility of the Father, lest man should at any time become a despiser of God, and that he should  always possess something towards which he might advance; but, on the other hand, revealing God to men through many dispensations, lest man, falling away from God altogether, should cease to exist. For the glory of God is a living man; and the life of man consists in beholding God. For if the manifestation of God which is made by means of the creation, affords life to all living in the earth, much more does that revelation of the Father which comes through the Word, give life to those who see God.

8. Inasmuch, then, as the Spirit of God pointed out by the prophets things to come, forming and adapting us beforehand for the purpose of our being made subject to God, but it was still a future thing that man, through the good pleasure of the Holy Spirit, should see [God], it necessarily behooved those through whose instrumentality future things were announced, to see God, whom they intimated as to be seen by men; in order that God, and the Son of God, and the Son, and the Father, should not only be prophetically announced, but that He should also be seen by all His members who are sanctified and instructed in the things of God, that man might be disciplined beforehand and previously exercised for a reception into that glory which shall afterwards be revealed in those who love God. For the prophets used not to prophesy in word alone, but in visions also, and in their mode of life, and in the actions which they performed, according to the suggestions of the Spirit. After this invisible manner, therefore, did they see God, as also Esaias says, ``I have seen with my eyes the King, the Lord of hosts,'' \gloss{Isaiah 6:5} pointing out that man should behold God with his eyes, and hear His voice. In this manner, therefore, did they also see the Son of God as a man conversant with men, while they prophesied what was to happen, saying that He who was not come as yet was present proclaiming also the impassible as subject to suffering, and declaring that He who was then in heaven had descended into the dust of death.  Moreover, [with regard to] the other arrangements concerning the summing up that He should make, some of these they beheld through visions, others they proclaimed by word, while others they indicated typically by means of [outward] action, seeing visibly those things which were to be seen; heralding by word of mouth those which should be heard; and performing by actual operation what should take place by action; but [at the same time] announcing all prophetically. Wherefore also Moses declared that God was indeed a consuming fire \gloss{Deuteronomy 4:24} (\textit{igneum}) to the people that transgressed the law, and threatened that God would bring upon them a day of fire; but to those who had the fear of God he said, ``The Lord God is merciful and gracious, and long-suffering, and of great commiseration, and true, and keeps justice and mercy for thousands, forgiving unrighteousness, and transgressions, and sins.'' \gloss{Exodus 34:6-7} 

9. And the Word spoke to Moses, appearing before him, ``just as any one might speak to his friend.'' \gloss{Numbers 12:8} But Moses desired to see Him openly who was speaking with him, and was thus addressed: ``Stand in the deep place of the rock, and with My hand I will cover you. But when My splendour shall pass by, then you shall see My back parts, but My face you shall not see: for no man sees My face, and shall live.'' \gloss{Exodus 33:20-22} Two facts are thus signified: that it is impossible for man to see God; and that, through the wisdom of God, man shall see Him in the last times, in the depth of a rock, that is, in His coming as a man. And for this reason did He [the Lord] confer with him face to face on the top of a mountain, Elias being also present, as the Gospel relates, \gloss{Matthew 17:3, etc.} He thus making good in the end the ancient promise.

10. The prophets, therefore, did not openly behold the actual face of God, but [they saw] the dispensations and the mysteries through which man should afterwards see God. As was also said to Elias: ``You shall go forth tomorrow, and stand in the presence of the Lord; and, behold, a wind great and strong, which shall rend the mountains, and break the rocks in pieces before the Lord. And the Lord [was] not in the wind; and after the wind an earthquake, but the Lord [was] not in the earthquake; and after the earthquake a fire, but the Lord [was] not in the fire; and after the fire a scarcely audible voice'' (\textit{vox auræ tenuis}). \gloss{1 Kings 19:11-12} For by such means was the prophet— very indignant, because of the transgression of the people and the slaughter of the prophets— both taught to act in a more gentle manner; and the Lord's advent as a man was pointed out, that it should be subsequent to that law which was given by Moses, mild and tranquil, in which He would neither break the bruised reed, nor quench the smoking flax. \gloss{Isaiah 42:3} The mild and peaceful repose of His kingdom was indicated likewise. For, after the wind which rends the mountains, and after the earthquake, and after the fire, come the tranquil and peaceful times of His kingdom, in which the spirit of God does, in the most gentle manner, vivify and increase mankind. This, too, was made still clearer by Ezekiel, that the prophets  saw the dispensations of God in part, but not actually God Himself. For when this man had seen the vision \gloss{Ezekiel 1:1} of God, and the cherubim, and their wheels, and when he had recounted the mystery of the whole of that progression, and had beheld the likeness of a throne above them, and upon the throne a likeness as of the figure of a man, and the things which were upon his loins as the figure of amber, and what was below like the sight of fire, and when he set forth all the rest of the vision of the thrones, lest any one might happen to think that in those [visions] he had actually seen God, he added: ``This was the appearance of the likeness of the glory of God.'' \gloss{Ezekiel 2:1} 

11. If, then, neither Moses, nor Elias, nor Ezekiel, who had all many celestial visions, saw God; but if what they did see were similitudes of the splendour of the Lord, and prophecies of things to come; it is manifest that the Father is indeed invisible, of whom also the Lord said, ``No man has seen God at any time.'' \gloss{John 1:18} But His Word, as He Himself willed it, and for the benefit of those who beheld, did show the Father's brightness, and explained His purposes (as also the Lord said: ``The only-begotten God,  which is in the bosom of the Father, He has declared [Him];'' and He does Himself also interpret the Word of the Father as being rich and great); not in one figure, nor in one character, did He appear to those seeing Him, but according to the reasons and effects aimed at in His dispensations, as it is written in Daniel. For at one time He was seen with those who were around Ananias, Azarias, Misaël, as present with them in the furnace of fire, in the burning, and preserving them from [the effects of] fire: ``And the appearance of the fourth,'' it is said, ``was like to the Son of God.'' \gloss{Daniel 3:26} At another time [He is represented as] ``a stone cut out of the mountain without hands,'' \gloss{Daniel 7:13-14} and as smiting all temporal kingdoms, and as blowing them away (\textit{ventilans ea}), and as Himself filling all the earth. Then, too, is this same individual beheld as the Son of man coming in the clouds of heaven, and drawing near to the Ancient of Days, and receiving from Him all power and glory, and a kingdom. ``His dominion,'' it is said, ``is an everlasting dominion, and His kingdom shall not perish.'' \gloss{Daniel 7:4} John also, the Lord's disciple, when beholding the sacerdotal and glorious advent of His kingdom, says in the Apocalypse: ``I turned to see the voice that spoke with me. And, being turned, I saw seven golden candlesticks; and in the midst of the candlesticks One like the Son of man, clothed with a garment reaching to the feet, and girt about the paps with a golden girdle; and His head and His hairs were white, as white as wool, and as snow; and His eyes were as a flame of fire; and His feet like fine brass, as if He burned in a furnace. And His voice [was] as the voice of waters; and He had in His right hand seven stars; and out of His mouth went a sharp two-edged sword; and His countenance was as the sun shining in his strength.'' \gloss{Revelation 1:12} For in these words He sets forth something of the glory [which He has received] from His Father, as [where He makes mention of] the head; something in reference to the priestly office also, as in the case of the long garment reaching to the feet. And this was the reason why Moses vested the high priest after this fashion. Something also alludes to the end [of all things], as [where He speaks of] the fine brass burning in the fire, which denotes the power of faith, and the continuing instant in prayer, because of the consuming fire which is to come at the end of time. But when John could not endure the sight (for he says, ``I fell at his feet as dead;'' \gloss{Revelation 1:17} that what was written might come to pass: ``No man sees God, and shall live'' \gloss{Exodus 33:20}), and the Word reviving him, and reminding him that it was He upon whose bosom he had leaned at supper, when he put the question as to who should betray Him, declared: ``I am the first and the last, and He who lives, and was dead, and behold I am alive for evermore, and have the keys of death and of hell.'' And after these things, seeing the same Lord in a second vision, he says: ``For I saw in the midst of the throne, and of the four living creatures, and in the midst of the elders, a Lamb standing as it had been slain, having seven horns, and seven eyes, which are the seven spirits of God, sent forth into all the earth.'' \gloss{Revelation 5:6} And again, he says, speaking of this very same Lamb: ``And behold a white horse; and He that sat upon him was called Faithful and True; and in righteousness does He judge and make war. And His eyes were as a flame of fire, and on His head were many crowns; having a name written, that no man knows but Himself: and He was girded around with a vesture sprinkled with blood: and His name is called The Word of God. And the armies of heaven followed Him upon white horses, clothed in pure white linen. And out of His mouth goes a sharp sword, that with it He may smite the nations; and He shall  rule (\textit{pascet}) them with a rod of iron: and He treads the wine-press of the fierceness of the wrath of God Almighty. And He has upon His vesture and upon His thigh a name written, King of Kings and Lord of Lords.'' \gloss{Revelation 19:11-17} Thus does the Word of God always preserve the outlines, as it were, of things to come, and points out to men the various forms (\textit{species}), as it were, of the dispensations of the Father, teaching us the things pertaining to God.

12. However, it was not by means of visions alone which were seen, and words which were proclaimed, but also in actual works, that He was beheld by the prophets, in order that through them He might prefigure and show forth future events beforehand. For this reason did Hosea the prophet take ``a wife of whoredoms,'' prophesying by means of the action, ``that in committing fornication the earth should fornicate from the Lord,'' \gloss{Hosea 1:2-3} that is, the men who are upon the earth; and from men of this stamp it will be God's good pleasure to take out \gloss{Acts 15:14} a Church which shall be sanctified by fellowship with His Son, just as that woman was sanctified by intercourse with the prophet. And for this reason, Paul declares that the ``unbelieving wife is sanctified by the believing husband.''  Then again, the prophet names his children, ``Not having obtained mercy,'' and ``Not a people,'' \gloss{Hosea 1:6-9} in order that, as says the apostle, ``what was not a people may become a people; and she who did not obtain mercy may obtain mercy. And it shall come to pass, that in the place where it was said, This is not a people, there shall they be called the children of the living God.'' \gloss{Romans 9:25-26} That which had been done typically through his actions by the prophet, the apostle proves to have been done truly by Christ in the Church. Thus, too, did Moses also take to wife an Ethiopian woman, whom he thus made an Israelitish one, showing by anticipation that the wild olive tree is grafted into the cultivated olive, and made to partake of its fatness. For as He who was born Christ according to the flesh, had indeed to be sought after by the people in order to be slain, but was to be set free in Egypt, that is, among the Gentiles, to sanctify those who were there in a state of infancy, from whom also He perfected His Church in that place (for Egypt was Gentile from the beginning, as was Ethiopia also); for this reason, by means of the marriage of Moses, was shown forth the marriage of the Word;  and by means of the Ethiopian bride, the Church taken from among the Gentiles was made manifest; and those who do detract from, accuse, and deride it, shall not be pure. For they shall be full of leprosy, and expelled from the camp of the righteous. Thus also did Rahab the harlot, while condemning herself, inasmuch as she was a Gentile, guilty of all sins, nevertheless receive the three spies,  who were spying out all the land, and hid them at her home; [which three were] doubtless [a type of] the Father and the Son, together with the Holy Spirit. And when the entire city in which she lived fell to ruins at the sounding of the seven trumpets, Rahab the harlot was preserved, when all was over [\textit{in ultimis}], together with all her house, through faith of the scarlet sign; as the Lord also declared to those who did not receive His advent,— the Pharisees, no doubt, nullify the sign of the scarlet thread, which meant the passover, and the redemption and exodus of the people from Egypt—when He said, ``The publicans and the harlots go into the kingdom of heaven before you.'' \gloss{Matthew 21:31} 
\end{document}
